---
title: "Simulating Data"
output: html_document
date: "2023-07-11"
---

# Models

The objective is to generate data for joint models. The longitudinal model is

```{=latex}
\begin{align*}
  Y_{ij} &= \beta_{11} + \beta_{12}t_{ij} + \beta_{13}X_i + U_{1i} + 
  U_{2i}t_{ij} + Z_{ij},
  \\
  X_i &\sim N(0,1),
  \\
  (U_{1i}, U_{2i}) &\sim MVN(0, \Sigma),
  \\
  Z_{ij} &\sim N(0, \sigma_z),
\end{align*}
```

for individual i and jth observation.

The model for event times is

```{=latex}
\begin{align*}
  h_i(t_i) &= h_{0i}(t_i) \exp\{\beta_{21}X_i + \gamma_1 U_{1i} + 
  \gamma_2 U_{2i} + \gamma_3(U_{1i} + U_{2i}t_i) + U_{3i}\},
  \\
  U_{3i} &\sim N(0, \sigma_3),
  \\
  h_{0i}(t_i) &= \lambda \rho t_i^{\rho-1} \text{(Weibull baseline)}.
\end{align*}
```

# Description

The following packages are needed:

```{r, message=FALSE, warning=FALSE}
library("MASS") 
library("cmdstanr")
library("pracma")
```


To generate the data we need to create $(U_1, U_2)$, $U_3$ and $X$:

```{r}
set.seed(123)

N<- 250
lambda <- 1
rho_s <- 2
cens_time <- 6
beta <- c(0,1,1,1)
gamma <- c(-1,1,1)
var_u <- c(1,1,1)
var_z <- 0.25
rho <- 0.5
n_rep_obs <- 0.5

beta_11 <- beta[1]
beta_12 <- beta[2]
beta_13 <- beta[3]
beta_21 <- beta[4]

gamma_1 <- gamma[1] 
gamma_2 <- gamma[2]
gamma_3 <- gamma[3]

mu_u1 <- 0
mu_u2 <- 0
mu_u <- c(mu_u1,mu_u2)
var_u1 <- var_u[1]
var_u2 <- var_u[2]
var_u3 <- var_u[3]
sigma <- matrix(c(var_u1, sqrt(var_u1*var_u2)*rho, 
                  sqrt(var_u1*var_u2)*rho, var_u2), 2)
bvn_u <- mvrnorm(N, mu = mu_u, Sigma = sigma)
u_1 <- bvn_u[,1]
u_2 <- bvn_u[,2]
u_3 <- rnorm(N, 0, var_u3)

x <- rnorm(N, 0, 1)
```

## Time-to-event data

For time-to-event data, we need to compute the survival probability 
$S_i(t) = \exp(-H_i(t))$, where $H_i(t) = \int_{0}^{t} h_i(s)ds$ is the 
cumulative hazard function. To generate $S_i(t)$, we can apply the procedure 
described in 
[this article](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.2059): we 
generate $V_i \sim U(0,1)$ and we take $t = S_i^{-1}(V_i)$. In R, we can create 
$H$ as a function of $t$ and find the root of 
$S_i(t) - V_i = \exp(-H_i(t)) - V_i = \exp(\int_{0}^{t} h_i(s)ds) - V_i$. We 
can try to integrate $h_i(s)$ using the integrate function and find the root 
using the uniroot function, but it doesn't work:

```{r, error=TRUE}
v <- runif(n=N)
times <- vector()

for(i in 1:N){
  haz <- Vectorize(function(s) {
    lambda*rho_s*s^(rho_s-1)*exp(beta_21*x[i] + gamma_1*u_1[i] + 
    gamma_2*u_2[i] + gamma_3*(u_1[i] + u_2[i]*s) + u_3[i])
  })
  
  cum_haz <- Vectorize(function(t) integrate(haz, 0, t)$value)
  
  sv <- Vectorize(function(t) exp(-cum_haz(t)) - v[i])
  
  times[i] <- uniroot(sv, lower = 0, upper = 200)$root
}
```

An alternative is calculate the result of the integral and use the lower 
incomplete gamma function, but it needs $-gamma_3*u_2[i]*u > 0$:

```{r, message=FALSE, warning=FALSE, error=TRUE}
for(i in 1:N){
  h_0_not_t <- lambda*rho_s
  h_not_t <- h_0_not_t*exp(beta_21*x[i] + gamma_1*u_1[i] + gamma_2*u_2[i] + 
             gamma_3*(u_1[i]) + u_3[i])
  
  ifelse(gamma_3*u_2[i] == 0, 
         cum_haz <- Vectorize(function(u) h_not_t*u^rho_s/rho_s),
         cum_haz <- Vectorize(function(u) h_not_t*(-gamma_3*u_2[i])^(-rho_s)*
                    pracma::gammainc(-gamma_3*u_2[i]*u, rho_s)[[1]]))
  
  sv <- Vectorize(function(u) exp(-cum_haz(u)) - v[i])
  
  times[i] <- uniroot(sv, lower = 0, upper = 200)$root
}
```

Replacing the uniroot function with the optim, the code work (the solution 
found minimizes the error of $|S_i - V_i|$ and may not be a root):

```{r, message=FALSE, warning=FALSE, error=TRUE}
for(i in 1:N){
  h_0_not_t <- lambda*rho_s
  h_not_t <- h_0_not_t*exp(beta_21*x[i] + gamma_1*u_1[i] + gamma_2*u_2[i] + 
             gamma_3*(u_1[i]) + u_3[i])
  
  ifelse(gamma_3*u_2[i] == 0, 
         cum_haz <- Vectorize(function(u) h_not_t*u^rho_s/rho_s),
         cum_haz <- Vectorize(function(u) h_not_t*(-gamma_3*u_2[i])^(-rho_s)*
                    pracma::gammainc(-gamma_3*u_2[i]*u, rho_s)[[1]]))
  
  sv <- Vectorize(function(u) abs(exp(-cum_haz(u)) - v[i]))
  
  times[i] <- optim(0, sv, lower = 0, upper = Inf, method = "L-BFGS-B")$par
}
```

However, the best solution found was to use the first code with the optim 
function:

```{r, error=TRUE}
for(i in 1:N){
  haz <- Vectorize(function(s) {
    lambda*rho_s*s^(rho_s-1)*exp(beta_21*x[i] + gamma_1*u_1[i] + 
    gamma_2*u_2[i] + gamma_3*(u_1[i] + u_2[i]*s) + u_3[i])
  })
  
  cum_haz <- Vectorize(function(t) integrate(haz, 0, t)$value)
  
  sv <- Vectorize(function(t) abs(exp(-cum_haz(t)) - v[i]))
  
  times[i] <- optim(0, sv, lower = 0, upper = Inf, method = "L-BFGS-B")$par
}
```

## Longitudinal data

For the longitudinal data, we create repeated observations up to time $t_i$:

```{r}
id <- long_out <- x_total <- obs_times_out <- vector()

for(i in 1:N){
  # number of repeated observations for each individual
  obs_times <- seq(0,times[i], by = n_rep_obs) 
  
  x_t <- rep(x[i], length(obs_times))

  x_total <- c(x_total,x_t)
  z = rnorm(length(obs_times), 0, var_z)
  yt <- beta_11 + beta_12*obs_times + beta_13*rep(x[i], length(obs_times)) + 
    rep(u_1[i], length(obs_times)) + rep(u_2[i], length(obs_times))*
    obs_times + z

  long_out <- c(long_out,yt)
  id <- c(id,rep(i,length(obs_times)))
  obs_times_out <- c(obs_times_out,obs_times)
}
```


# Functions

Generating data:

```{r}
sim_data <- function(N, 
                     lambda, 
                     rho_s, 
                     cens_time, 
                     beta, 
                     gamma, 
                     var_u, 
                     var_z, 
                     rho, 
                     n_rep_obs){
  
  times <- id <- long_out <- x_total <- vector()
  
  beta_11 <- beta[1]
  beta_12 <- beta[2]
  beta_13 <- beta[3]
  beta_21 <- beta[4]
  
  gamma_1 <- gamma[1] 
  gamma_2 <- gamma[2]
  gamma_3 <- gamma[3]
  
  mu_u1 <- 0
  mu_u2 <- 0
  mu_u <- c(mu_u1,mu_u2)
  var_u1 <- var_u[1]
  var_u2 <- var_u[2]
  var_u3 <- var_u[3]
  sigma <- matrix(c(var_u1, sqrt(var_u1*var_u2)*rho, 
                    sqrt(var_u1*var_u2)*rho, var_u2), 2)
  bvn_u <- mvrnorm(N, mu = mu_u, Sigma = sigma)
  u_1 <- bvn_u[,1]
  u_2 <- bvn_u[,2]
  u_3 <- rnorm(N, 0, var_u3)
  
  x <- rnorm(N, 0, 1)
  
  ###################
  # Survival process
  ###################
  
  # Simulating the times to event

  v <- runif(n=N)
  id_times <- c(1:N)
  
  for(i in 1:N){
    haz <- Vectorize(function(s) {
      lambda*rho_s*s^(rho_s-1)*exp(beta_21*x[i] + gamma_1*u_1[i] + 
      gamma_2*u_2[i] + gamma_3*(u_1[i] + u_2[i]*s) + u_3[i])
    })
    
    cum_haz <- Vectorize(function(t) integrate(haz, 0, t)$value)

    sv <- Vectorize(function(t) abs(exp(-cum_haz(t)) - v[i]))

    times[i] <- optim(1e-6, sv, lower = 0, upper = Inf, method = "L-BFGS-B")$par
  }
  
  status <- as.vector(times < cens_time)
  times <- as.vector(ifelse(status, times, cens_time))
  status <- as.numeric(status) # Censoring indicators (1=Observed, 0=Censored)
  
  ##############################
  # Longitudinal process  
  ##############################
  
  obs_times_out <- vector()
  for(i in 1:N){
    # number of repeated observations for each individual
    obs_times <- seq(0,times[i], by = n_rep_obs) 
    
    x_t <- rep(x[i], length(obs_times))

    x_total <- c(x_total,x_t)
    z = rnorm(length(obs_times), 0, sqrt(var_z))
    y_t <- beta_11 + beta_12*obs_times + beta_13*rep(x[i], length(obs_times)) + 
          rep(u_1[i], length(obs_times)) + rep(u_2[i], length(obs_times))*
          obs_times + z

    long_out <- c(long_out,y_t)
    id <- c(id,rep(i,length(obs_times)))
    obs_times_out <- c(obs_times_out,obs_times)
  }
  
  #---------------------------------------------------------------------
  # Creating the longitudinal and survival processes object
  #---------------------------------------------------------------------
  long_proc <- as.matrix(cbind(id, long_out, x_total, obs_times_out)) 
  surv_proc <- as.matrix(cbind(id_times, x, times, status)) 
  data <- list(long_proc, surv_proc)
  names(data) <- c("longitudinal","survival")
  
  return(data)
}
```


Creating data:

```{r}
N<- 100
lambda <- 1
rho_s <- 2
cens_time <- 6
beta <- c(0,1,1,1)
gamma <- c(0,0,0)
var_u <- c(0.5,0.5,0.5)
var_z <- 0.25
rho <- 0.5
n_rep_obs <- 0.5

data <- sim_data(N, 
                 lambda, 
                 rho_s, 
                 cens_time, 
                 beta, 
                 gamma, 
                 var_u, 
                 var_z, 
                 rho, 
                 n_rep_obs)
```

Stan model for longitudinal process:

```{stan output.var="long_model"}
functions{
  /**
  * Return a vector corresponding to linear predictor for longitudinal
  * sub-model.
  *
  * @param x Matrix of covariates
  * @param obs_times Vector of observed times
  * @param id Vector of integer values corrensponding to identifiers for each
           subject
  * @param beta_1 Vector corresponding to fixed effects
  * @param u Vector corresponding random effects.
  * @return Linear predictor vector.
  */
  vector linear_predictor(matrix x, 
                          vector obs_times, 
                          array[] int id, 
                          vector beta_1, 
                          matrix u){
    int N = num_elements(obs_times);
    vector[N] out;
    
    out = beta_1[1] + beta_1[2]*obs_times + beta_1[3]*x[id,1] + u[id,1] + 
          rows_dot_product(u[id,2],obs_times);
    
    return out;
  } 
}


data{
  int n_obs;
  int N;
  vector[n_obs] y;
  matrix[N,1] x;
  array[n_obs] int<lower=1,upper=N> id;
  vector[n_obs] obs_times;
}


parameters{
  vector[3] beta_1;
  real<lower=0> var_z;
  array[2] real<lower=0> var_u;
  real<lower=-1, upper=1> rho;
  matrix[N,2] u;
}

transformed parameters{
  cov_matrix[2] sigma;

  sigma[1,1] = var_u[1];
  sigma[2,2] = var_u[2];
  sigma[1,2] = sqrt(var_u[1]*var_u[2])*rho;
  sigma[2,1] = sigma[1,2];
}


model{
  // ------------------------------------------------------
  //        LOG-LIKELIHOOD FOR LONGITUDINAL SUBMODEL                
  // ------------------------------------------------------
  
  vector[n_obs] linpred; 

  // Linear predictor
  linpred = linear_predictor(x, obs_times, id, beta_1, u);

  // Longitudinal Normal log-likelihood
  target += normal_lpdf(y | linpred, sqrt(var_z));

  // ------------------------------------------------------
  //                       LOG-PRIORS                       
  // ------------------------------------------------------
    
  // Longitudinal fixed effects
  target += normal_lpdf(beta_1 | 0, 100);
  
  // Random effects
  for(i in 1:N){ 
    target += multi_normal_lpdf(u[i,1:2] | rep_vector(0.0,2), sigma);
  }

  // Random effects variance
  target += inv_gamma_lpdf(var_u | 0.01, 0.01);

  // Random effects correlation
  target += beta_lpdf((rho+1)/2 | 0.5, 0.5);
  
  // Residual error variance
  target += inv_gamma_lpdf(var_z | 0.01, 0.01);
}
```

Stan model for survival process:

```{stan output.var="event_model"}
functions{
  /**
  * Return a real number corresponding to function function to be integrated 
  * evaluated at point t.
  *
  * @param t Value to evaluate function  
  * @param xc A high precision version of the distance from x to the nearest 
  *        endpoint in a definite integral
  * @param theta Parameter values used to evaluate the integral
  * @param x_r Data values used to evaluate the integral
  * @param x_i Integer data used to evaluate the integral.
  * @return The value of the integrand evaluated at the point x.
  */
  real integrand(real t, 
                 real xc, 
                 array[] real theta, 
                 array[] real x_r, 
                 array[] int x_i) {
    real rho_s = theta[1];
    real gamma_3 = theta[2];
    real u_2i = theta[3];
    
    real h = t^(rho_s-1)*exp(gamma_3*u_2i*t);
      
    return h;    
  }
  
  /**
  * Return a real number corresponding to function to be integrated 
  * evaluated at point t.
  *
  * @param t Value to evaluate function  
  * @param xc A high precision version of the distance from x to the nearest 
  *        endpoint in a definite integral
  * @param theta Parameter values used to evaluate the integral
  * @param x_r Data values used to evaluate the integral
  * @param x_i Integer data used to evaluate the integral.
  * @return The value of the integrand evaluated at the point x.
  */
  real integrand_bp(real t,
                    real xc,                
                    array[] real theta,     
                    array[] real x_r,                        
                    array[] int x_i) {
    real rho_s = theta[1];
    real gamma_3 = theta[2];
    real u_2i = theta[3];
    
    real h = t^rho_s*exp(gamma_3*u_2i*t);
      
    return h;    
  }

  /**
  * Return a Laplace approximation of specific function.
  *
  * @param a Lower limit to evaluate integral  
  * @param b Upper limit to evaluate integral
  * @param rho_s Shape of the Weibull baseline hazard
  * @param gamma_3 3th association parameter
  * @param u_2i ith value of the second random effect.
  * @return The integral approximation.
  */
  real lap_app(real a, real b, real rho_s, real gamma_3, real u_2i){
    real t0 = (1-rho_s)/(gamma_3*u_2i);
    real h_t0 = (rho_s-1)*log(t0) + gamma_3*u_2i*t0;
    real ddh_t0 = (1-rho_s)/(t0^2);
    real lout = h_t0 + 0.5*(log(2) + log(pi()) - log(-ddh_t0)) +
                log(normal_cdf(b, t0, -ddh_t0) -
                normal_cdf(a, t0, -ddh_t0));
    // real out = exp(h_t0)*sqrt(2*pi()/(-ddh_t0))*
    //            (normal_cdf(b, t0, -ddh_t0) -
    //            normal_cdf(a, t0, -ddh_t0));
    return exp(lout);
  }

  /**
  * Return a lower incomplete gamma function aproximation.
  *
  * @param a Integrand parameter  
  * @param z Upper limit to evaluate integral
  * @param tol Aproximation tolerance
  * @return The approximation.
  */
  real low_inc_gamma(real a, real z, real tol) {
    real inc_gamma;
    
    if (z < -50) {
      int n = 0;
      real inc_gamma0 = (1-a)/(z^n);
      real inc_gamma1 = inc_gamma0 + 1;
      while (abs(inc_gamma1 - inc_gamma0) > tol){
        n += 1;
        inc_gamma1 = inc_gamma0;
        inc_gamma0 += (1-a)/(z^n);
      }
      inc_gamma = exp(z)/(z*tgamma(a))*inc_gamma0;
    }
    else {
      int n = 0;
      real inc_gamma0 = (-z)^n/(tgamma(n+1)*(a+n));
      real inc_gamma1 = inc_gamma0 + 1;
      while (abs(inc_gamma1 - inc_gamma0) > tol){
        n += 1;
        inc_gamma1 = inc_gamma0;
        inc_gamma0 += (-z)^n/(tgamma(n+1)*(a+n));
      }
      inc_gamma = 1/(tgamma(a))*inc_gamma0;
    }
    real out = z^a*tgamma(a)*inc_gamma;
    return out;
  }
  
  /**
  * Return a vector with each value out[i] corresponding to logarithm hazard 
  * function evaluated at point t[i].
  *
  * @param t Vector corresponding to observed time values  
  * @param x Matrix of covariates
  * @param u_1 Vector corresponding to random effects
  * @param u_2 Vector corresponding to random effects
  * @param u_3 Vector corresponding to random effects
  * @param gamma Vector corresponding to association parameters
  * @param beta_21 Value corresponding to fixed effects
  * @param lambda Scale of the Weibull baseline hazard
  * @param rho_s Shape of the Weibull baseline hazard
  * @param id Array corresponding to subjects to evaluate the function.
  * @return Vector corresponding to logarithm hazard function evaluated at 
  * observed times.
  */
  vector log_haz(vector t, 
                matrix x, 
                vector u_1, 
                vector u_2, 
                vector u_3, 
                vector gamma, 
                real beta_21, 
                real lambda, 
                real rho_s,
                array[] int id){
    int N = num_elements(id);
    vector[N] out;
    real gamma_1 = gamma[1];
    real gamma_2 = gamma[2];
    real gamma_3 = gamma[3];
    out = rep_vector(log(lambda), N) + rep_vector(log(rho_s), N) + (rho_s-1)*
          log(t[id]) + beta_21*x[id,1] + gamma_1*u_1[id] + gamma_2*u_2[id] +
          gamma_3*(u_1[id] + rows_dot_product(u_2[id],t[id])) + u_3[id];
    return out;
  }

  /**
  * Return a vector with each value out[i] corresponding to constant terms of
  * the cumulative hazard function evaluated at point t[i]
  *
  * @param t Vector corresponding to observed time values  
  * @param x Matrix of covariates
  * @param u_1 Vector corresponding to random effects
  * @param u_2 Vector corresponding to random effects
  * @param u_3 Vector corresponding to random effects
  * @param gamma Vector corresponding to association parameters
  * @param beta_21 Value corresponding to fixed effects
  * @param lambda Scale of the Weibull baseline hazard
  * @param rho_s Shape of the Weibull baseline hazard.
  * @return Vector corresponding to constant terms of the cumulative hazard 
  * function evaluated at observed times.
  */
  vector const_term_cum_haz(vector t, 
                           matrix x, 
                           vector u_1, 
                           vector u_2, 
                           vector u_3, 
                           vector gamma, 
                           real beta_21, 
                           real lambda, 
                           real rho_s){
    int N = num_elements(t);                         
    vector[N] out;
    real gamma_1 = gamma[1];
    real gamma_2 = gamma[2];
    real gamma_3 = gamma[3];
    out = lambda*rho_s*exp(beta_21*x[1:N,1] + gamma_1*u_1 + 
          gamma_2*u_2 + gamma_3*u_1 + u_3);
    return out;
  }
  
  /**
  * Return a vector with each value out[i] corresponding to log constant terms 
  * of the cumulative hazard function evaluated at point t[i]
  *
  * @param t Vector corresponding to observed time values  
  * @param x Matrix of covariates
  * @param u_1 Vector corresponding to random effects
  * @param u_2 Vector corresponding to random effects
  * @param u_3 Vector corresponding to random effects
  * @param gamma Vector corresponding to association parameters
  * @param beta_21 Value corresponding to fixed effects
  * @param lambda Scale of the Weibull baseline hazard
  * @param rho_s Shape of the Weibull baseline hazard.
  * @return Vector corresponding to log constant terms of the cumulative hazard 
  * function evaluated at observed times.
  */
    
  vector lconst_term_cum_haz(vector t, 
                             matrix x, 
                             vector u_1, 
                             vector u_2, 
                             vector u_3, 
                             vector gamma, 
                             real beta_21, 
                             real lambda, 
                             real rho_s){
    int N = num_elements(t);
    vector[N] out;
    real gamma_1 = gamma[1];
    real gamma_2 = gamma[2];
    real gamma_3 = gamma[3];
    out = rep_vector(log(lambda), N) + rep_vector(log(rho_s), N) +
           beta_21*x[1:N,1] + gamma_1*u_1 + gamma_2*u_2 + gamma_3*u_1 + u_3;
    return out;
  }
}


data{
  int N;
  int n_unc_times;
  matrix[N,1] x;
  vector[N] times;
  array[n_unc_times] int<lower=1,upper=N> ind_unc_times;
}

transformed data {
  array[0] real x_r;
  array[0] int x_i;
}

parameters{
  real beta_21;
  vector[3] gamma;
  real<lower=0> lambda;
  real<lower=0> rho_s;
  array[2] real<lower=0> var_u;
  real<lower=-1, upper=1> rho;
  real<lower=0> var_u3;
  matrix[N,2] u;
  vector[N] u_3;
}

transformed parameters{
  cov_matrix[2] sigma;

  sigma[1,1] = var_u[1];
  sigma[2,2] = var_u[2];
  sigma[1,2] = sqrt(var_u[1]*var_u[2])*rho;
  sigma[2,1] = sigma[1,2];
}


model {
  // ------------------------------------------------------
  //          LOG-LIKELIHOOD FOR SURVIVAL SUBMODEL                
  // ------------------------------------------------------
  
  vector[n_unc_times] lhaz;

  vector[N] lsurv;
  
  // Log-hazard function
  lhaz = log_haz(times,
                x,
                u[1:N,1],
                u[1:N,2],
                u_3,
                gamma,
                beta_21,
                lambda,
                rho_s,
                ind_unc_times);
                
  // Integration by parts

  // Log-survival function
  lsurv = -const_term_cum_haz(times,
                             x,
                             u[1:N,1],
                             u[1:N,2],
                             u_3,
                             gamma,
                             beta_21,
                             lambda,
                             rho_s);

  for (i in 1:N) {
    real integral = integrate_1d(integrand_bp,
                                 0.0,
                                 times[i],
                                 {rho_s, gamma[3], u[i,2]},
                                 x_r,
                                 x_i,
                                 1e-8);
    lsurv[i] *= (1/rho_s)*(times[i]^rho_s*exp(gamma[3]*u[i,2]*times[i]) -
                gamma[3]*u[i,2]);
  }

  // // Integration by parts with log
  // 
  // // Log-survival function
  // lsurv = lconst_term_cum_haz(times,
  //                             x,
  //                             u[1:N,1],
  //                             u[1:N,2],
  //                             u_3,
  //                             gamma,
  //                             beta_21,
  //                             lambda,
  //                             rho_s);
  // 
  // for (i in 1:N) {
  //   real integral = integrate_1d(integrand_bp,
  //                                0.0,
  //                                times[i],
  //                                {rho_s, gamma[3], u[i,2]},
  //                                x_r,
  //                                x_i,
  //                                0.01);
  //   lsurv[i] += -log(rho_s) +
  //               log((times[i]^rho_s*exp(gamma[3]*u[i,2]*times[i]) -
  //               gamma[3]*u[i,2]*integral));
  //   lsurv[i] = -exp(lsurv[i]);
  // }

  // Integration using incomplete gamma approximation
  // for (i in 1:N) {
  //   if (!is_nan(low_inc_gamma(rho_s, -gamma[3]*u[i,2]*times[i], 0.01))){
  //     lsurv[i] *= (-gamma[3]*u[i,2])^(-rho_s)*
  //                 low_inc_gamma(rho_s, -gamma[3]*u[i,2]*times[i], 0.01);
  //   }
  //   else {
  //     real integral = integrate_1d(integrand_bp,
  //                                  0.0,
  //                                  times[i],
  //                                  {rho_s, gamma[3], u[i,2]},
  //                                  x_r,
  //                                  x_i,
  //                                  0.01);
  //     lsurv[i] *= (1/rho_s)*
  //                 (times[i]^rho_s*exp(gamma[3]*u[i,2]*times[i]) -
  //                 gamma[3]*u[i,2]*integral);
  //   }
  // }

  // Direct integral
  // for (i in 1:N) {
  //   real integral = integrate_1d(integrand,
  //                                0.0,
  //                                times[i],
  //                                {rho_s, gamma[3], u[i,2]},
  //                                x_r,
  //                                x_i,
  //                                0.01);
  //   lsurv[i] = lsurv[i]*integral;
  // }
  
  // // Laplace Approximation
  // for (i in 1:N) {
  //   if (rho_s > 1){
  //     lsurv[i] *= lap_app(0, times[i], rho_s, gamma[3], u[i,2]);
  //   }
  //   else {
  //     real integral = integrate_1d(integrand_bp,
  //                                  0.0,
  //                                  times[i],
  //                                  {rho_s, gamma[3], u[i,2]},
  //                                  x_r,
  //                                  x_i,
  //                                  0.01);
  //     lsurv[i] *= (1/rho_s)*
  //                 (times[i]^rho_s*exp(gamma[3]*u[i,2]*times[i]) -
  //                 gamma[3]*u[i,2]*integral);
  //   }
  // }
   
  // Survival log-likelihood
  target += sum(lhaz) + sum(lsurv);
   
   
  // ------------------------------------------------------
  //          PRIOR DISTRIBUTIONS
  // ------------------------------------------------------

  // Survival fixed effects
  target += normal_lpdf(beta_21 | 1, 5);

  // Weibull scale parameter
  target += gamma_lpdf(lambda | 1, 1);

  // Weibull shape parameter
  target += gamma_lpdf(rho_s | 2, 1);

  // Association parameters
  target += normal_lpdf(gamma | 0, 10);

  // Random effects
  for(i in 1:N){
    target += multi_normal_lpdf(u[i,1:2] | rep_vector(0.0,2), sigma);
    target += normal_lpdf(u_3[i] | 0, var_u3);
  }

  // Random effects variance
  target += inv_gamma_lpdf(var_u | 0.01, 0.01);
  target += inv_gamma_lpdf(var_u3 | 0.01, 0.01);

  // Random effects correlation
  target += beta_lpdf((rho+1)/2 | 2, 1);
}

```


# Results

Longitudinal model:

```{r}
setwd(paste("C:/Users/ezequ/OneDrive - Fundacao Getulio Vargas - FGV/Mentoria/",
            "joint-models-in-stan", sep = ""))

# Required quantities for longitudinal model fitting

x <- as.matrix(data$survival[,2],1) # unique x
N <- size(x)[1]                     # total number of observations
y <- data$longitudinal[,2]          # longitudinal outcomes
n_obs <- length(y)                  # total number of longitudinal outcomes
id <- data$longitudinal[,1]         # patient IDs
obs_times <- data$longitudinal[,4]  # visit times for repeated observations

long_data = list(y=y,
                 N=N,
                 n_obs=n_obs,
                 x=x,
                 id=id,
                 obs_times=obs_times)

long_model <- cmdstan_model("code/long_model.stan")

long_posterior_samples <- long_model$sample(data = long_data, chains = 1)
```

```{r}
long_posterior_samples$summary()
```

Event-time model:

```{r}
setwd(paste("C:/Users/ezequ/OneDrive - Fundacao Getulio Vargas - FGV/Mentoria/",
            "joint-models-in-stan", sep = ""))

# Required quantities for event-time model fitting

x <- as.matrix(data$survival[,2],1)  # unique x
N <- size(x)[1]                      # total number of observations
status <- data$survival[,4]          # vital status (1 = dead, 0 = alive)
times <- data$survival[,3]           # times to event
ind_unc_times <- which(status==1)    # uncensored times indicator
n_unc_times <- length(ind_unc_times) # number of uncensored times

event_data <- list(N=N,
                   x=x,
                   times=times,
                   ind_unc_times=ind_unc_times,
                   n_unc_times=n_unc_times)

event_model <- cmdstan_model("code/event_model.stan")

event_posterior_samples <- event_model$sample(data = event_data, chains = 1)
```

```{r}
event_posterior_samples$summary()
```


Codes based on:

- [How to create a toy survival (time to event) data with right censoring](https://stats.stackexchange.com/questions/135124/how-to-create-a-toy-survival-time-to-event-data-with-right-censoring)
- [A tractable Bayesian joint model for longitudinal and survival data](https://github.com/daniloalvares/Tractable-BJM/tree/main)




